Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts

### 2017-12-03 article/article.tex
adiantada sintese de textos

\subsection{Syntax}
\subsubsection{Fundamental notions}
As in other natural languages, 
colloquial Toki Pona might have
incomplete sentences and deviate from the
norm.
The basic structure of sentences is:
`subject' (Noun) li `predicate' (Verb) e `object' (Noun).
The li might be repeated to associate more than
one predicate to the subject.
The particle li is omitted if the subject is a simple mi (I or us)
or sina (you). A discussion about issues (problems)
yield by this rule
and how I deal with them is in Appendix~\ref{mytoki}.

@@

\subsection{Statistics of the vocabulary}\label{sec:stat}
In~\cite{pije} are statistics about Toki Pona corpus.
This section focuses on the statistics of the vocabulary
and syntactic rules:
the letters, phonemes, word sizes,
possible combinations for words and sentences.
The statistics are in Appendix~\cite{listings},
and the next paragraph is an overview.
Python scripts were used to obtain the measurements
and are available at~\cite{tokipona}.

@@

From the official 124 words, 
26 of them (20.97\%)
have only one syllable,
85 (68.55\%) have two syllables,
and 13 (10.48\%) have three syllables.
No official words has four or more syllables.
Of all the 235 syllables of the official words,
li, la, ka, na, and pa are the most often (with 5.53\%,
4.26\%, 3.83\%,  3.83\%, 3.83\%).
On the first syllable of each word,
a, o, pi, ka, and la are the most often
(with 6.45\%, 4.03\%, 4.03\%, 3.23\%, 3.23\%).
On the last syllable of each word,
li, lo, na, la, and ma are the most often
(with 8.06\%, 4.84\%, 4.84\%, 4.03\%, 4.03\%).
Middle-word syllables only occur 13 times,
and are all different, with the exception of `la',
which occurs twice.
For the frequencies of all syllables in every position,
run \tttt{makeStatistics.py} in~\cite{tokipona}.

@@

Given the rules given in Section~\ref{phonology}
(of 14 letters, 5 vowels, (C)V(N) phonemes,
forbidden ji, wu, wo, ti, nn, nm),
46 words are possible with 1 syllable,
1886 with 2 syllables, 77326 with 3 syllables.
No middle syllable end with the nasal consonant 'n'.
These are shown in Table~\ref{tab:pw}
with fraction of the space actually used in Toki Pona.

Possible syntactic structures are quantified
in Table~\ref{tab:syn} for
Possibilities on noun, verb and prepositional phrases
and in sentences.
(Counting along the number of words.)

A toolbox for obtaining such statistics and sets
of words is in~\cite{tokipona}.

% number of possible syllables
% number of possible words of size X (1, 2, 3, 4....).

\subsection{Synthesis of text}
Such counting exercise is also useful
for (semi-)automate writing through scripting.
The syntax organizes the words in larger structures.
The rhymes are restricted,
the ending is as a tonality in music.

There are some specific tasks for achieving texts,
such as finding the number of syllables considering the elisoes,
or handling interaction of the writer with the script
to choose sentences or verses or stanzas.

The same package~\cite{tokipona}
has capabilities for synthesizing text.
Noun, verb and prepositional phrases,
sentences.
It also aims at making larger scale texts
by keeping a record of the used words and structures (context)
and using stylistic outlines for poems and short narratives.

Figure~\ref{fig:syn} holds some synthesized texts
with different coloring schemes for syntax highlighting,
as presented in the next section.

@@

\section{Conclusions and further work}\label{conc}
\begin{itemize}
  \item Make a syntax highlighting that colors the tokens with respect
    to the syntatic position and function, and not in a fixed manner
    as is implemented and exposed in Section~\ref{shigh}.
  \item Relate Toki Pona to Wordnet: should one Toki Pona word
    be related to more then one synset of the English language?
    An immediate implementation would be linking each Toki Pona word
    to the English words in the official vocabulary.
    A Toki Pona synset would be related to all the synsets of each English
    words regarded as lemmas (sequence of characters or sound).
  \item Translators. This Toki Pona Wordnet would be useful e.g. for naive translators
    of words and text.
    Enhancements of such translators would use POS tagging and templates
    in English and Toki Pona and/or corpus and further statistical methods.
    An efficiency of Toki Pona can be studied in terms of the coverage of
    the Wordnet cloud.
  \item Conlanging. In fact, I plan to use Wordnet for making new conlangs
    (conlanging), and insights derived from Toki Pona might be used for that end.
  \item Understand how the corpus is gathered in \url{tokipona.net}.
  \item Know about previously existing words that were used for Toki Pona
    (e.g. suno and suwi might come from sun and sweet),
    and about the reasons that lead Sonja (and maybe other people)
    to choose the 14 letters and the syllable structure.
    This might require a dedicated communication with the
    speaker community and the documentation authors.
  \item Corpus-based analysis.
  \item Publication of original texts and translations.
  \item Make an article written in Toki Pona.
    I wrote Section~\ref{finalToki}, and believe that
    a summary in both English and Toki Pona for
    facilitating the acquisition of context,
    a reasonable article on some scientific topics is
    possible.
    I first conceived something around complexity, statistics, physics, or computer science.
    But one possibility is to write about linguistics, philosophy, literature
    or psychology with the partners I write in English.
    I can start a draft,
    qhey might learn the language in a few hours (with or without my help),
    to contribute, and we can write a short paper.
  \item Enhance the synthesis of text to yield better contextualized text
    and stylistic traces, such as for poems and short stories.
    Give the user the ability to choose the sentences (generate randomly according to
    previously written or given text,
    some rules input by the user, the package and the language guidelines,
    outputs to the screen and asks to keep and discard).
  \item Enhance the syntax highlighting implementation described in Section~\ref{high}.
    It uses only syntactic cues to choose POS tags,
    which might be overcome by using n-grams and
    further techniques from Natural Language Processing~\cite{POS}.
  \item We are particularly interested in software tools,
    and one possibility is to use a conlang to describe software
    routines.
    In this sense, it seems profitable
    to understand what `non-ambiguous' means in Lojban,
    and in what sense one is able to compile and parse Lojban.
    This study will probably yield considerations about
    parsing Toki Pona files.
  \item Conlang idealized for describing data, programs,
    scientific writings, creative writing
    (exploring the thinking process).
    Different modes that might be specified by a section header or tag.
\end{itemize}

@@

\begin{thebibliography}{99}
\fontsize{11}{0}\selectfont
\bibitem{anPh}
	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
	Available at \url{https://youtu.be/oeOKYc3-nbM}
\bibitem{anPh2}
	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What_are_you_and_I_anthropological_physics_fundamentals_}
\bibitem{ouroWiki}
  Ouroboros. (2017, November 2). In Wikipedia, The Free Encyclopedia. Retrieved 22:19, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ouroboros&oldid=808392809}
\bibitem{vimrc}
	Fabbri, R. (2017). A reasonable vimrc file. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/vimrc} 
\bibitem{ex}
  Ex (text editor). (2017, March 22). In Wikipedia, The Free Encyclopedia. Retrieved 22:22, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ex_(text_editor)&oldid=771621020}
\bibitem{tokipona}
	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
\bibitem{tpLang}
	Lang, S. (2014). Toki Pona: the language of good. Tawhid Publishing.
    ISBN-10: 0978292308, ISBN-13: 978-0978292300.
\bibitem{aa1}
	Fabbri, R., Fabbri, R., Vieira, V., Penalva, D., Shiga, D., Mendonça, M., Negrao, A., Zambianchi, L., \& Thumé, G. (2013). AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology. RESI. From \url{https://arxiv.org/abs/1604.08255}
\bibitem{aa2}
	Fabbri, R. (2017).
The Algorithmic-Autoregulation (AA) Methodology and Software:
a collective focus on self-transparency. ENMC2017. From \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf} 
\bibitem{kama}
  Knight, B. (2017) o kama sona e toki pona! Available at:
    \url{http://tokipona.net/tp/janpije/okamasona.php}
\bibitem{Wikipesija}
  Toki Pona community (2017). Wikipesija. Available at:
    \url{http://tokipona.wikia.com}
\bibitem{conlanWikip}
  Constructed language. (2017, November 1). In Wikipedia, The Free Encyclopedia. Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
\end{thebibliography}
\end{document}

>>>>

\subsection{Syntax}\label{syntax}
\subsubsection{Fundamental notions}
As in other natural languages, 
colloquial Toki Pona might have
incomplete sentences and deviate from the
norm.
The basic structure of sentences is:
`subject' (Noun) li `predicate' (Verb) e `object' (Noun).
The li might be repeated to associate more than
one predicate to the subject.
The particle li is omitted if the subject is a simple mi (I or us)
or sina (you). A discussion about issues (problems)
yield by this rule
and how I deal with them is in Appendix~\ref{mytoki}.

@@

The Python scripts (or the toolbox)
for obtaining all the results in this sections,
and more, is publicly available in~\cite{tokipona},
to facilitate both the inspection of the results and the
generation of derivatives by other interested parties.

\subsection{Statistics of the vocabulary}\label{sec:stat}
In~\cite{pije} are statistics about Toki Pona corpus.
This section focuses on the statistics of the vocabulary
and syntactic rules:
the letters, phonemes, word sizes,
possible combinations for words and sentences.
The statistics are in Appendix~\cite{listings},
and the next paragraph is an overview.
The script \tttt{makeStatistics.py}
was used to obtain the all the measurements and tables
discussed herein,
and hold some other measurements and sets of words
which were regarded as less suitable for this exposition.

@@

From the official 124 words, 
26 of them (20.97\%)
have only one syllable,
85 (68.55\%) have two syllables,
and 13 (10.48\%) have three syllables.
No official words has four or more syllables.
Of all the 235 syllables of the official words,
li, la, ka, na, and pa are the most often (with 5.53\%,
4.26\%, 3.83\%,  3.83\%, 3.83\%).
On the first syllable of each word,
a, o, pi, ka, and la are the most often
(with 6.45\%, 4.03\%, 4.03\%, 3.23\%, 3.23\%).
On the last syllable of each word,
li, lo, na, la, and ma are the most often
(with 8.06\%, 4.84\%, 4.84\%, 4.03\%, 4.03\%).
Middle-word syllables only occur 13 times,
and are all different, with the exception of `la',
which occurs twice.
For the frequencies of all syllables in every position,
run the script \tttt{makeStatistics.py} in~\cite{tokipona}.

@@

Given the rules given in Section~\ref{phonology}
(of 14 letters, 5 vowels, (C)V(N) phonemes,
forbidden ji, wu, wo, ti, nn, nm),
96 words are possible with 1 syllable,
8256 with 2 syllables, 710016 with 3 syllables.
No middle syllable end with the nasal consonant 'n'.

Possible syntactic structures are unlimited.
For a noun or verb phrase, one might quantify the possibilities
by considering all words but the particles that are nothing else
(i.e. li, e, la, pi, a, o, anu, en, seme, mu).
This yields $120-10=110 words$.
As discussed in Section~\ref{syntax},
I advocate that, although such words are classified in the dictionary,
they might be used indistinctly as nouns, adjectives,
verbs and adverbs.
Thus, one have 110 possibilities of noun and verb phrases with
one word, $110^2=12100$ possibilities with two words,
$110^3=1331000$ with three words and so on.
Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
(verb), object (noun) and prepositional (noun or verb) phrases of a sentence,
and assume that the sentence has at most one prepositional phrase,
one might quantify the possibilities using the formula:
$\delta = 110^n\times 110^v\times 10^o \times 5\times 110^p$,
where $5$ stands for the possible prepositions.
To account for the particles, one possibility is
to assume for simplicity that one might use none or one particle
(not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
at each phrase, yielding:
$\delta = 110^n\times 110^v\times 110^o \times 5\times 110^p\times 9^4$.
For example, assume $n=v=p=o=1$, then
$\delta=110^1\times 110^1\times 110^1 \times 5\times 110^1\times 9^4=
4.80298^{12}$
i.e. almost 5 trillion possible
sentences with single word phrases (i.e. no adjectives or adverbs),
while allowing only one particle per phrase
and only one predicate phrase 
(e.g. 'ona li moku e soweli lon supa').
The general case is implemented as a function
in the script \tttt{makeStatistics.py}~\cite{tokipona}

\subsection{Synthesis of text}\label{synth}
Such counting exercises is also useful
for (semi-)automated writing through scripting.
The syntax organizes the words in larger structures.
The rhymes are restricted and sonorities are
restricted by the small vocabulary and simple syntax.
There are some specific tasks for achieving texts,
such as finding the number of syllables considering the elisions,
or handling interaction of the writer with the script
to choose sentences or verses or stanzas.

The same package~\cite{tokipona}
has capabilities for synthesizing Toki Pona text.
Noun, verb and prepositional phrases,
sentences.
It also aims at making larger scale texts
by keeping a record of the used words and structures (context)
and using stylistic outlines for poems and short narratives.

Figure~\ref{fig:syn} holds some synthesized texts
with different coloring schemes for syntax highlighting,
as presented in the next section.
The textual synthesis described here,
and implemented in \tttt{makeTexts.py}
might be enhanced as described in Section~\ref{conc}.

@@

\section{Conclusions and further work}\label{conc}
\begin{itemize}
  \item Further develop the text synthesis facilities described in
    Section~\ref{synth} might be enhanced by further employing e.g.
    rhythm, meter, rhyme and form techniques~\cite{wikipPoetry}.
  \item Make a syntax highlighting that colors the tokens with respect
    to the syntatic position and function, and not in a fixed manner
    as is implemented and exposed in Section~\ref{shigh}.
  \item Relate Toki Pona to Wordnet: should one Toki Pona word
    be related to more then one synset of the English language?
    An immediate implementation would be linking each Toki Pona word
    to the English words in the official vocabulary.
    A Toki Pona synset would be related to all the synsets of each English
    words regarded as lemmas (sequence of characters or sound).
  \item Translators. This Toki Pona Wordnet would be useful e.g. for naive translators
    of words and text.
    Enhancements of such translators would use POS tagging and templates
    in English and Toki Pona and/or corpus and further statistical methods.
    An efficiency of Toki Pona can be studied in terms of the coverage of
    the Wordnet cloud.
  \item Conlanging. In fact, I plan to use Wordnet for making new conlangs
    (conlanging), and insights derived from Toki Pona might be used for that end.
  \item Understand how the corpus is gathered in \url{tokipona.net}.
  \item Know about previously existing words that were used for Toki Pona
    (e.g. suno and suwi might come from sun and sweet),
    and about the reasons that lead Sonja (and maybe other people)
    to choose the 14 letters and the syllable structure.
    This might require a dedicated communication with the
    speaker community and the documentation authors.
  \item Corpus-based analysis.
  \item Publication of original texts and translations.
  \item Make an article written in Toki Pona.
    I wrote Section~\ref{finalToki}, and believe that
    a summary in both English and Toki Pona for
    facilitating the acquisition of context,
    a reasonable article on some scientific topics is
    possible.
    I first conceived something around complexity, statistics, physics, or computer science.
    But one possibility is to write about linguistics, philosophy, literature
    or psychology with the partners I write in English.
    I can start a draft,
    qhey might learn the language in a few hours (with or without my help),
    to contribute, and we can write a short paper.
  \item Enhance the synthesis of text to yield better contextualized text
    and stylistic traces, such as for poems and short stories.
    Give the user the ability to choose the sentences (generate randomly according to
    previously written or given text,
    some rules input by the user, the package and the language guidelines,
    outputs to the screen and asks to keep and discard).
  \item Enhance the syntax highlighting implementation described in Section~\ref{high}.
    It uses only syntactic cues to choose POS tags,
    which might be overcome by using n-grams and
    further techniques from Natural Language Processing~\cite{POS}.
  \item We are particularly interested in software tools,
    and one possibility is to use a conlang to describe software
    routines.
    In this sense, it seems profitable
    to understand what `non-ambiguous' means in Lojban,
    and in what sense one is able to compile and parse Lojban.
    This study will probably yield considerations about
    parsing Toki Pona files.
  \item Conlang idealized for describing data, programs,
    scientific writings, creative writing
    (exploring the thinking process).
    Different modes that might be specified by a section header or tag.
\end{itemize}

@@

\begin{thebibliography}{99}
\fontsize{11}{0}\selectfont
\bibitem{anPh}
	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
	Available at \url{https://youtu.be/oeOKYc3-nbM}
\bibitem{anPh2}
	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What_are_you_and_I_anthropological_physics_fundamentals_}
\bibitem{ouroWiki}
  Ouroboros. (2017, November 2). In Wikipedia, The Free Encyclopedia. Retrieved 22:19, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ouroboros&oldid=808392809}
\bibitem{vimrc}
	Fabbri, R. (2017). A reasonable vimrc file. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/vimrc} 
\bibitem{ex}
  Ex (text editor). (2017, March 22). In Wikipedia, The Free Encyclopedia. Retrieved 22:22, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ex_(text_editor)&oldid=771621020}
\bibitem{tokipona}
	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
\bibitem{tpLang}
	Lang, S. (2014). Toki Pona: the language of good. Tawhid Publishing.
    ISBN-10: 0978292308, ISBN-13: 978-0978292300.
\bibitem{aa1}
	Fabbri, R., Fabbri, R., Vieira, V., Penalva, D., Shiga, D., Mendonça, M., Negrao, A., Zambianchi, L., \& Thumé, G. (2013). AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology. RESI. From \url{https://arxiv.org/abs/1604.08255}
\bibitem{aa2}
	Fabbri, R. (2017).
The Algorithmic-Autoregulation (AA) Methodology and Software:
a collective focus on self-transparency. ENMC2017. From \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf} 
\bibitem{kama}
  Knight, B. (2017) o kama sona e toki pona! Available at:
    \url{http://tokipona.net/tp/janpije/okamasona.php}
\bibitem{Wikipesija}
  Toki Pona community (2017). Wikipesija. Available at:
    \url{http://tokipona.wikia.com}
\bibitem{conlanWikip}
  Constructed language. (2017, November 1). In Wikipedia, The Free Encyclopedia. Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
\bibitem{wikiPoetry}
Poetry. (2017, November 19). In Wikipedia, The Free Encyclopedia.
    Retrieved 19:46, December 3, 2017, from
    \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
\end{thebibliography}
\end{document}
